Hydra battles in Coq

### 2020-10-04 doc/hydras.tex
minor modifications

\subsubsection{A predicate for characterizing normal forms}
\label{sec:orgheadline74}
\label{sec:orgheadline75}
Our data-type \texttt{T1} allows us to write expressions that
are not properly in Cantor normal form as specified in Section \ref{sec:epsilon0-intro}.
For instance, consider the following term of type  \texttt{T1}.

@@

The type \texttt{T1} is not a correct ordinal notation, since it contains terms that are not in Cantor normal form. Moreover,
in Sections~\ref{sect:L-equations}, \ref{sect:hardy},
and \ref{sect:wainer}, we needed to define total recursive functions on well-formed ordinal terms less  than $\epsilon_0$, using the \texttt{Equations} plug-in~\cite{sozeau:hal-01671777}.
 In order to define a type whose inhabitants represent just ordinals, we build a type gathering a term of type \texttt{T1} and a proof that this term is in normal form.

@@

\index{Exercises}
\begin{exercise}
In earlier versions of this development, the predicate \texttt{nf} was defined  inductively, with various constructors describing all possible cases.
\begin{enumerate}
\item Please give such a definition, in a dedicated module.
\item Prove the logical equivalence between your definition and ours.
\item Define a variant of the type \texttt{E0} (with your definition of \texttt{nf}).
\item Can you still prove a lemma like \texttt{E0\_eq\_iff} ? With the help of an axiom of the standard library ?
\end{enumerate}
\end{exercise}
For upgrading constants and fonctions of \texttt{T1}, we have to prove that 
the term they build is in normal form.
For instance, let us represent the ordinals $0$ and $\omega$ as instances of the class \texttt{E0}.

@@

\begin{Coqsrc}
Lemma plus_not_monotonous : exists alpha beta gamma : T1,
    alpha t1< beta /\ alpha + gamma = beta + gamma.
Proof.
  exists 3, 5, omega;  now  compute.
Qed.

@@

Let us consider again the ordinal $\alpha_0$ defined in section~\vref{alpha0-def}

If we ask \coq{} to print the normal form of \texttt{alpha\_0}, we get a hardly readable term of type \texttt{T1}.

@@

The following data type defines a abstract syntax fore more readable ordinals in Cantor normal form:

@@

\subsection{About  well-foundedness}
\label{sec:orgheadline82}
   In order to use \texttt{T1} for proving termination results,
we need to prove that  our representation of ordinals strictly less than \(\epsilon_0\)
makes our order \texttt{<} well-founded. Then we will get \emph{transfinite induction} for free.

The proof of well-foundedness of the strict order $<$ on Cantor normal forms is already 
available in the Cantor contribution by Castéran and Contéjean~\cite{CantorContrib}. That proof relies on a library on recursive path orderings written by
E. Contéjean. We present also a direct proof of the same result, which does not require any knowledge on r.p.o.s.

@@

The problem comes from that \(\delta\) may be bigger that \(\beta\) or
\(\gamma\);
for instance \(\delta\) may be of the form
\texttt{ocons $\beta'$ $p'$  $\gamma'$},
where  \(\beta' \leq  \beta\) and  \(p' < n\).
Thus, the induction hypotheses \texttt{IHbeta} and \texttt{IHgamma}  are useless for finishing our proof.

@@

The following lemma is an application of the strict inequality 
\showmath{\alpha < \omega ^\alpha}. If \showmath{\omega^\alpha} is accessible,
then \showmath{\alpha} is \emph{a fortori} accessible.

@@

First, we prove that, for  any \texttt{LT}-accessible term \showmath{\alpha}, any well formed
term (\texttt{ocons $\alpha$ $n$ $\beta$})  is accessible too.

@@

We built an instance of \texttt{ON}, and proved its correction w.r.t. Schutte's model.

@@

\emph{Coq}'s constraints on recursive definitions resulted in 
the quite  complex form of \texttt{oplus}'s definition.
For making easier proof of properties of this function, 
it is helpful to \emph{derive} lemmas that will help to simplify 
expressions of the form (\texttt{oplus $a$ $ b$}).

A first set of lemmas correspond to the various cases of \texttt{oplus}'s 
definition. They can be proved almost immediately, using \texttt{cbn} 
and \texttt{rewrite} tactics.

@@

\begin{Coqsrc}
  Lemma oplus_comm : forall alpha beta, 
      nf alpha -> nf beta ->
      alpha o+ beta =  beta o+ alpha.
  Proof with eauto with T1.
    intros alpha beta Halpha Hbeta;
    apply oplus_comm_0 with (succ (max alpha beta)) ...  
  (* rest of proof omitted *)
\end{Coqsrc}

@@

\begin{Coqsrc}
 Lemma oplus_assoc_0 :
    forall alpha,
      nf alpha ->
      forall a b c,  nf a -> nf b -> nf c ->
                      lt a alpha ->
                      lt b alpha -> lt c alpha ->
                      a o+ (b o+ c) = (a o+ b) o+ c.
  Proof with eauto with T1.
    intros alpha Halpha.
    transfinite_induction alpha.
(* rest of proof omitted *) 
\end{Coqsrc}

\begin{Coqsrc}
 Lemma oplus_assoc : forall alpha beta gamma,
                        nf alpha -> nf beta -> nf gamma ->
                                    alpha o+ (beta o+ gamma) =
                                    alpha o+ beta o+ gamma.
 Proof with eauto with T1.
    intros;
    apply oplus_assoc_0 with (succ (max alpha (max beta gamma))) ...
(* rest of proof omitted *)   
\end{Coqsrc}

@@

\subsection{A measure for hydra-battle termination}
\label{sec:orgheadline90}
\label{sec:hydra-measure}

@@

@@

In their article~\cite{KP82}, Kirby and Paris show that this result cannot be proved in Peano arithmetic. Their proof uses some knowledge about model theory and non-standard models of Peano arithmetic. In this chapter, we focus on a specific class of proofs of termination of hydra battles: construction of some variant mapping the type \texttt{Hydra} into some segment of ordinals. Our proof relies only on the Calculus of Inductive Constructions and is a natural complement of the results proven in the previous chapter.

\begin{itemize}
\item There is no variant mapping the type \texttt{Hydra} into the interval $[0,\omega^2)$ (section ~\vref{omega2-case}), and a fortiori 
$[0,\omega)$

\item There is a variant that maps the type \texttt{Hydra} into the
interval $[0,\epsilon_0)$ (theorem \texttt{every\_battle\_terminates}, in section~\vref{hydra-variant}).
\end{itemize}

@@

Canonical sequences are functions that associate an ordinal $\canonseq{\alpha}{i}$ to every ordinal $\alpha<\epsilon_0$ and positive integer $i$. They satisfy nice properties :

@@

\item If $\beta<\alpha<\epsilon_0$, then there is a ``path'' from $\alpha$ to $\beta$, \emph{i.e.} a
sequence $\alpha_0=\alpha, \alpha_1, \dots, \alpha_n=\beta$, where for every $k<n$, there exists some $i_k$ such that $\alpha_{k+1}=\canonseq{\alpha_k}{i_k}$
\item Canonical sequences correspond tightly to rounds of hydra battles: if $\alpha\not=0$,
then $\iota(\alpha)$ is transformed into $\iota(\canonseq{\alpha}{i+1}$ in one round with
the replication factor $i$ (Lemma \href{../src/html/hydras.Hydra.O2H.html\#canonS_iota_i}{Hydra.O2H.canonS\_iota\_i}).
\item From the two previous properties, we infer that whenever $\beta<\alpha<\epsilon_0$, there exists a (free) battle from $\iota(\alpha)$ to $\iota(\beta)$.
\end{itemize}

\begin{remark}
  In~\cite{KS81}, canonical sequences are defined for any ordinal $\alpha <\epsilon_0$,
by stating that if $\alpha$ is a successor ordinal $\beta+1$,  the sequence associated with 
$\alpha$ is simply the constant sequence whose terms are equal to $\beta$.
Likewise, we define the canonical sequence of $0$ as the 
sequence all terms of which are equal to $0$.

>>>>

\subsubsection{A predicate for characterizing normal forms}
\label{sect:t1-nf}

\label{sec:orgheadline74}
\label{sec:orgheadline75}
Our data-type \texttt{T1} allows us to write expressions that
are not properly in Cantor normal form as specified in Section \ref{sec:epsilon0-intro}.
For instance, consider the following term of type  \texttt{T1}.

@@

As we noticed in Sect.~\ref{sect:t1-nf}, the type \texttt{T1} is not a correct ordinal notation, since it contains terms that are not in Cantor normal form. In certain contexts (for instance in Sections~\ref{sect:L-equations}, \ref{sect:hardy},
and \ref{sect:wainer}),  we need to define total recursive functions on well-formed ordinal terms less  than $\epsilon_0$, using the \texttt{Equations} plug-in~\cite{sozeau:hal-01671777}.
 In order to define a type whose inhabitants represent just ordinals, we build a type gathering a term of type \texttt{T1} and a proof that this term is in normal form.

@@

\index{Exercises}
\begin{exercise}
In earlier versions of this development, the predicate \texttt{nf} was defined  inductively, with various constructors describing all possible cases.
\begin{enumerate}
\item Please give such a definition, in a dedicated module.
\item Prove the logical equivalence between your definition and ours.
\item Define a variant of the type \texttt{E0} (with your definition of \texttt{nf}).
\item Can you still prove a lemma like \texttt{E0\_eq\_iff} ? With the help of an axiom from some module of the standard library ?
\end{enumerate}
\end{exercise}
For upgrading constants and fonctions of \texttt{T1}, we have to prove that 
the term they build is in normal form.
For instance, let us represent the ordinals $0$ and $\omega$ as instances of the class \texttt{E0}.

@@

\begin{Coqsrc}
Lemma plus_not_monotonous_l : exists alpha beta gamma : T1,
    alpha t1< beta /\ alpha + gamma = beta + gamma.
Proof.
  exists 3, 5, omega;  now  compute.
Qed.

@@

Let us consider again the ordinal $\alpha_0$ defined in section~\vref{alpha0-def}
If we ask \coq{} to print its  normal form, we get a hardly readable term of type \texttt{T1}.

@@

The following data type defines a abstract syntax for more readable ordinals in Cantor normal form:

@@

\subsection{About  well-foundedness}
\label{sec:orgheadline82}
   In order to use \texttt{T1} for proving termination results,
we need to prove that  our order \texttt{<} is well-founded. Then we will get \emph{transfinite induction} for free.

The proof of well-foundedness of the strict order $<$ on Cantor normal forms is already 
available in the Cantor contribution by Castéran and Contejean~\cite{CantorContrib}. That proof relies on a library on recursive path orderings written by
E. Contejean. We present here  a direct proof of the same result, which does not require any knowledge on r.p.o.s.

@@

The problem comes from the hypothesis \texttt{Hdelta}. It does not prevent  \(\delta\) to be bigger that \(\beta\) or
\(\gamma\);
for instance \(\delta\) may be of the form
\texttt{ocons $\beta'$ $p'$  $\gamma'$},
where  \(\beta' \leq  \beta\) and  \(p' < n\).
Thus, the induction hypotheses \texttt{IHbeta} and \texttt{IHgamma}  are useless for finishing our proof.

@@

The following lemma is an application of the strict inequality 
\showmath{\alpha < \omega ^\alpha}. If \showmath{\omega^\alpha} is accessible,
then \showmath{\alpha} is \emph{a fortiori} accessible.

@@

First, we prove that, for  any \texttt{LT}-accessible term \showmath{\alpha}, \showmath{\alpha} is 
strongly accessible too (\emph{i.e.} any well formed
term (\texttt{ocons $\alpha$ $n$ $\beta$})  is accessible).

@@

We build an instance of \texttt{ON}, and prove its correction w.r.t. Schutte's model.

@@

\emph{Coq}'s constraints on recursive definitions result in 
the quite  complex form of \texttt{oplus}'s definition.
Proofs of properties of this function can be simpler if we
 \emph{derive} rewriting lemmas that will help to simplify 
expressions of the form (\texttt{oplus $a$ $ b$}).

A first set of lemmas correspond to the various cases of \texttt{oplus}'s 
definition. They can be proved almost immediately, using \texttt{cbn} 
and \texttt{reflexivity} tactics.

@@

\begin{Coqsrc}
  Lemma oplus_comm : forall alpha beta, 
      nf alpha -> nf beta ->
      alpha o+ beta =  beta o+ alpha.
  Proof with eauto with T1.
    intros alpha beta Halpha Hbeta;
    apply oplus_comm_0 with (succ (max alpha beta)) ...  
  (* ... *)
\end{Coqsrc}

@@

\begin{Coqsrc}
 Lemma oplus_assoc_0 :
    forall alpha,
      nf alpha ->
      forall a b c,  nf a -> nf b -> nf c ->
                      lt a alpha ->
                      lt b alpha -> lt c alpha ->
                      a o+ (b o+ c) = (a o+ b) o+ c.
  Proof with eauto with T1.
    intros alpha Halpha.
    transfinite_induction alpha.
    (* ... *)
\end{Coqsrc}

\begin{Coqsrc}
 Lemma oplus_assoc : forall alpha beta gamma,
                        nf alpha -> nf beta -> nf gamma ->
                                    alpha o+ (beta o+ gamma) =
                                    alpha o+ beta o+ gamma.
 Proof with eauto with T1.
    intros;
    apply oplus_assoc_0 with (succ (max alpha (max beta gamma))) ...
    (* ... *)   
\end{Coqsrc}

@@

\subsection{A measure for hydra-battle termination}

\label{sec:hydra-measure}

@@

\label{thm:every-battle-terminates}

@@

In their article~\cite{KP82}, Kirby and Paris show that this result cannot be proved in Peano arithmetic. Their proof uses some knowledge about model theory and non-standard models of Peano arithmetic. In this chapter, we focus on a specific class of proofs of termination of hydra battles: construction of some variant mapping the type \texttt{Hydra} into a given initial  segment of ordinals. Our proof relies only on the Calculus of Inductive Constructions and is a natural complement of the results proven in the previous chapter.

\begin{itemize}
\item There is no variant mapping the type \texttt{Hydra} into the interval $[0,\omega^2)$ (section ~\vref{omega2-case}), and a fortiori 
$[0,\omega)$ (section ~\vref{omega-case}).

\item There exists a variant which maps the type \texttt{Hydra} into the
interval $[0,\epsilon_0)$ (theorem \texttt{every\_battle\_terminates}, in section~\vref{thm:every-battle-terminates}).
\end{itemize}

@@

Canonical sequences are functions that associate an ordinal $\canonseq{\alpha}{i}$ to every ordinal $\alpha<\epsilon_0$ and positive integer $i$. They satisfy several nice properties:

@@

\item If $\beta<\alpha<\epsilon_0$, then there is a ``path'' from $\alpha$ to $\beta$, \emph{i.e.} a
sequence $\alpha_0=\alpha, \alpha_1, \dots, \alpha_n=\beta$, where for every $k<n$, there exists some $i_k$ such that $\alpha_{k+1}=\canonseq{\alpha_k}{i_k}$
\item Canonical sequences correspond tightly to rounds of hydra battles: if $\alpha\not=0$,
then $\iota(\alpha)$ is transformed into $\iota(\canonseq{\alpha}{i+1})$ in one round with
the replication factor $i$ (Lemma \href{../src/html/hydras.Hydra.O2H.html\#canonS_iota_i}{Hydra.O2H.canonS\_iota\_i}).
\item From the two previous properties, we infer that whenever $\beta<\alpha<\epsilon_0$, there exists a (free) battle from $\iota(\alpha)$ to $\iota(\beta)$.
\end{itemize}

\begin{remark}
  In~\cite{KS81}, canonical sequences are defined for any ordinal $\alpha <\epsilon_0$,
by stating that if $\alpha$ is a successor ordinal $\beta+1$,  the sequence associated with 
$\alpha$ is simply the constant sequence whose terms are equal to $\beta$.
Likewise, the canonical sequence of $0$ maps any natural number to $0$.
